---
title: "Data Visualization Homework 4"
author: "Mineh Nazarian"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(janitor)
library(reshape2)
library(ggthemes)
library(viridis)
library(gridExtra)

# Load and clean data
bundesliga <- read_csv("bundesliga.csv") %>%
  clean_names()

# Manually create 'ftr' column
bundesliga <- bundesliga %>%
  mutate(ftr = case_when(
    fthg > ftag ~ "H",
    fthg < ftag ~ "A",
    fthg == ftag ~ "D",
    TRUE ~ "-"
  ))

bundesliga2 <- read_csv("bundesliga2.csv") %>%
  clean_names()
```


```{r setup, include=FALSE}
# Create FTR manually since I was getting errors

bundesliga <- read_csv("bundesliga.csv") %>% clean_names()
bundesliga2 <- read_csv("bundesliga2.csv") %>% clean_names()
```

# Part 1: Trend Analysis

## 1. Total and Average Goals Per Season

```{r}
goal_stats <- bundesliga %>%
  group_by(season) %>%
  summarise(
    total_goals = sum(fthg + ftag, na.rm = TRUE),
    matches = n(),
    avg_goals_per_match = total_goals / matches
  )

ggplot(goal_stats, aes(x = season)) +
  geom_line(aes(y = total_goals), color = "blue", size = 1.2) +
  geom_line(aes(y = avg_goals_per_match * 100), color = "red", size = 1.2) +
  labs(title = "Total and Average Goals per Season",
       subtitle = "Red line scaled x100 for average goals per match",
       y = "Goals / Avg * 100", x = "Season")
```

**Interpretation:**  
This chart shows total goals in blue and average goals per match in red (scaled ×100) for better comparison. You can observe how scoring intensity changed over the years.

## 2. Goal Distribution Per Match

```{r}
bundesliga <- bundesliga %>%
  mutate(goals = fthg + ftag)

season_avg <- bundesliga %>%
  group_by(season) %>%
  summarise(avg_goals = mean(goals, na.rm = TRUE))

bundesliga <- bundesliga %>%
  left_join(season_avg, by = "season") %>%
  mutate(over_under = ifelse(avg_goals > 2.5, "Over", "Under"))

ggplot(bundesliga, aes(x = goals, fill = over_under)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.7) +
  facet_wrap(~season, scales = "free_y") +
  labs(title = "Goal Distribution per Match by Season",
       fill = "Avg > 2.5 Goals") +
  theme_minimal()
```

**Interpretation:**  
This shows the distribution of goals per match each season, color-coded by whether the average exceeded 2.5 goals.

## 3. Team Goal Trends per Season (Highlight Bayern Munchen)

```{r}
pdf("season_goal_trends.pdf", width = 10, height = 6)

unique_seasons <- unique(bundesliga$season)

for (s in unique_seasons) {
  season_data <- bundesliga %>%
    filter(season == s) %>%
    group_by(team = hometeam) %>%
    summarise(goals = sum(fthg, na.rm = TRUE))

  gg <- ggplot(season_data, aes(x = team, y = goals, group = 1)) +
    geom_line(color = "grey") +
    geom_point(data = subset(season_data, team == "Bayern Munchen"),
               aes(x = team, y = goals), color = "red", size = 3) +
    labs(title = paste("Goals Scored by Teams in Season", s),
         subtitle = paste("Total goals:", sum(season_data$goals)),
         caption = paste("Bayern Munchen goals:", season_data$goals[season_data$team == "Bayern Munchen"])) +
    theme(axis.text.x = element_text(angle = 90))
  print(gg)
}

dev.off()
```

**Interpretation:**  
This saves a PDF file with per-season goal trends for all teams, highlighting Bayern Munchen in red.

## 4. Heatmap of Home vs. Away Wins

```{r}
heatmap_data <- bundesliga %>%
  filter(ftr %in% c("H", "A")) %>%
  mutate(result = ifelse(ftr == "H", hometeam, awayteam)) %>%
  group_by(season, result) %>%
  summarise(wins = n()) %>%
  pivot_wider(names_from = season, values_from = wins, values_fill = 0)

heatmap_matrix <- as.matrix(heatmap_data[,-1])
rownames(heatmap_matrix) <- heatmap_data$result

heatmap(heatmap_matrix, scale = "none", col = viridis(256), margins = c(5,10))
```

**Interpretation:**  
Heatmap showing how many home/away wins each team had across seasons.


# Part 5: Overall Performance (R - Total Points)

```{r}
bundesliga <- bundesliga %>%
  mutate(
    home_points = case_when(
      ftr == "H" ~ 3,
      ftr == "D" ~ 1,
      TRUE ~ 0
    ),
    away_points = case_when(
      ftr == "A" ~ 3,
      ftr == "D" ~ 1,
      TRUE ~ 0
    )
  )

home_points_df <- bundesliga %>%
  group_by(season, team = hometeam) %>%
  summarise(points = sum(home_points, na.rm = TRUE), .groups = "drop")

away_points_df <- bundesliga %>%
  group_by(season, team = awayteam) %>%
  summarise(points = sum(away_points, na.rm = TRUE), .groups = "drop")

total_points <- bind_rows(home_points_df, away_points_df) %>%
  group_by(season, team) %>%
  summarise(total_points = sum(points), .groups = "drop")

unique_teams <- unique(total_points$team)
team_colors <- setNames(rainbow(length(unique_teams)), unique_teams)

pdf("r_total_points_per_team.pdf", width = 10, height = 6)
for (s in unique(total_points$season)) {
  season_data <- total_points %>% filter(season == s) %>% arrange(total_points)
  bar_colors <- team_colors[season_data$team]
  barplot(
    season_data$total_points,
    names.arg = season_data$team,
    las = 2,
    col = bar_colors,
    main = paste("Total Points -", s),
    ylab = "Points"
  )
}
dev.off()
```

# Part 6: Monte Carlo Simulation (R)

```{r}
set.seed(42)

simulate_goals_r <- function(mean_goals, seasons = 10, reps = 1000) {
  replicate(reps, sum(rpois(seasons, lambda = mean_goals)))
}

team_avg_goals <- bundesliga %>%
  filter(hometeam %in% c("Bayern Munich", "Dortmund", "Leverkusen")) %>%
  group_by(hometeam) %>%
  summarise(mean_goals = mean(fthg, na.rm = TRUE))

sim_results <- lapply(team_avg_goals$mean_goals, simulate_goals_r)
names(sim_results) <- team_avg_goals$hometeam

boxplot(
  sim_results,
  names = names(sim_results),
  main = "Monte Carlo Simulation: Next 10 Seasons",
  ylab = "Total Simulated Goals",
  col = c("red", "blue", "green")
)



```

